---
title: "Control structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Control structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Control structures in R{.tabset .tabset-pills}


1. If-else statement

2. for cycles

3. while 

4. Repeat



## If-else statement 

Conditional execution of a code. General form:
```{r}
# if (<condition>){
# what to do if true
#} else if (<other condition>){
# what to do if true
#} else {
# what to do if none of them are true
#}
#
```

- Condition should be logical, or (0,1). 
- Else and if else are optional, multiple else if are possible.
- The statement will be executed and the first! expression that is true will be evaluated. 
- If multiple statements can be true and we want to evaluate all -> nested if statements.
- Short version is `ifelse()`.


## For cycle


Iteration, repeat behaviour. 
If we want to repeat the same command on a set of values or variables.  

```{r}
# for (i in 1:10) {
# print(i)
# }


```

- The value of the iterating variable (here `i`) is automatically changing in every cycle. 
- The iterating variable can be also logical, character, list element. 
- The return value of the iterating variable is the cycle that was run last. 


## While


While statements will repeat a task until a condition is met. 
Can easily be infinite, so be careful. 


```{r}

# count <- 0
# while(count < 10) {
#         print(count)
#         count <- count + 1
# }

##DO NOT RUN##

# count <- 0
# while(count < 10) {
#         print(count)
#         count <- count - 1
# }

```


## Repeat

`Repeat` will repeat the expression in an infinite loop. 
To stop, the `break` command has to be used. Can be dangerous.


```{r}

#count <- 1
# repeat {
#         count <- count + 1
# if (count >10)
#break
# }

```

## break and next

As seen above, `break` breaks the whole loop and exits. 

`next` only breaks the current iteration and the loop continues. 

```{r}

# for (i in 1:10) {
# if (i==5)
# next
# print(i)
# }

```

# Troubleshooting

[**Bingo!**](https://docs.google.com/presentation/d/1nSdvByo3jtu47sXF6lybmoGbPxh5xLp8o9aeRC-Bh20/edit)

Useful functions: print(), browser()

```{r}

 for (i in 1:10) {
   a <- i*2
  #browser()
   print(i)
 }

```


# Special looping functions{.tabset .tabset-pills}


Widely used functions: apply, lapply, sapply, tapply

## `apply()` 

  works on 2D objects (data.frame, matrix)
  
  iterates around rows or columns
  
  returns vectors

```{r}
dt <- data.frame(mes1=c(2,3,4), mes2=c(3,2,1), mes3=c(2,4,6))
rowmean <- apply(dt, 1, mean)
rowmean
colmean <- apply(dt, 2, mean)
colmean
```


## `lapply()` 


  Works on lists
  
  Iterates around list elements
  
  Returns a list
  

```{r}
lt <- list(c(1,2,4), c(4,52,5,3,2), c(2,3,5,2,2))
lapply(lt, length)
```


## `tapply()` 


  Works on subsettable objects (e.g. vector)
  
  Subsets them by the index and applies a function 
  
  Returns a list
  
  One can use `aggregate()` instead
  

```{r}
data(iris)
tapply(iris$Sepal.Width, iris$Species, median)
```


## `sapply()`, `vapply()`


  Similar to lapply
  
  Returns a lists, matrices, vectors or arrays. 


```{r}

i39 <- sapply(3:9, seq) # list of vectors
sapply(i39, fivenum)
vapply(i39, fivenum,
       c(Min. = 0, "1st Qu." = 0, Median = 0, "3rd Qu." = 0, Max. = 0))

```



