---
title: "Visualisation with ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualisation with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# VISUALISATION WITH ggplot2


Tutorial adapted from:
https://rafalab.github.io/dsbook/ggplot2.html

Load the __tidyverse__ libraries

```{r}
library(tidyverse)
```

## The components of a graph

We will construct a graph that summarizes the US murders dataset that looks like this:

```{r ggplot-example-plot, echo=FALSE}
library(dslabs)
data(murders)
library(ggthemes)
library(ggrepel)

# r <- murders %>% 
#   summarize(pop=sum(population), tot=sum(total)) %>% 
#   mutate(rate = tot/pop*10^6) %>% pull(rate)

murders %>% ggplot(aes(x = population/10^6, y = total, label = abb)) +  
  # geom_abline(intercept = log10(r), lty=2, col="darkgrey") +
  geom_point(aes(color=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name="Region") +
  theme_economist()
```


This data visualization shows us a lot of information. The code needed to make this plot is relatively simple. We will learn to create the plot part by part. 

The first step in learning __ggplot2__ is to be able to break a graph apart into components. Let's break down the plot above and introduce some of the __ggplot2__ terminology. The main three components to note are:
 
* __Data__: The US murders data table is being summarized. We refer to this as the __data__ component. 
*  __Geometry__: The plot above is a scatterplot. This is referred to as the 
__geometry__ component. 
* __Aesthetic mapping__: The plot uses several visual cues to represent the information provided by the dataset. The two most important cues in this plot are the point positions on the x-axis and y-axis, which represent population size and the total number of murders, respectively. Each point represents a different observation, and we _map_ data about these observations to visual cues like x- and y-scale. Color is another visual cue that we map to region. We refer to this as the __aesthetic mapping__ component. How we define the mapping depends on what __geometry__ we are using. 

We also note that:

* The points are labeled with the state abbreviations.
* The range of the x-axis and y-axis appears to be defined by the range of the data. They are both on log-scales. 
* There are labels, a title, a legend, and we use the style of The Economist magazine.

We will now construct the plot piece by piece.

We start by loading the dataset:

```{r}
library(dslabs)
data(murders)
```

## `ggplot` objects 
The first step in creating a __ggplot2__ graph is to define a `ggplot` object. We do this with the function `ggplot`, which initializes the graph. If we read the help file for this function, we see that the first argument is used to specify what data is associated with this object: 

We can  pipe the data in as the first argument. S

```{r ggplot-example-2}
murders %>% ggplot()
```

It renders a plot, in this case a blank slate since no geometry has been defined. The only style choice we see is a white background.


## Geometries

In `ggplot2` we create graphs by adding _layers_. Layers can define geometries, compute summary statistics, define what scales to use, or even change styles.
To add layers, we use the symbol `+`. In general, a line of code will look like this:

>> DATA %>% `ggplot()` + LAYER 1 + LAYER 2 + ... + LAYER N

The function used to create plots with this geometry is `geom_point`. 


## Aesthetic mappings
 
  __Aesthetic mappings__ describe how properties of the data connect with features of the graph, such as distance along an axis, size, or color.
  
```{r, eval = FALSE}
murders %>% ggplot() + 
  geom_point(aes(x = population/10^6, y = total))
```
 
We can drop the `x = ` and `y =` if we wanted to since these are the first and second expected arguments, as seen in the help page. 

Instead of defining our plot from scratch, we can also add a layer to the `p` object that is defined  as `p <- ggplot(data = murders)`:

```{r ggplot-example-3}
p <- ggplot(data = murders)
p + geom_point(aes(population/10^6, total))
```


The scale and labels are defined by default when adding this layer. 

## Layers

A second layer in the plot we wish to make involves adding a label to each point to identify the state. The `geom_label` and `geom_text` functions permit us to add text to the plot with and without a rectangle behind the text, respectively.

Because each point (each state in this case) has a label, we need an aesthetic mapping to make the connection between points and labels. 

```{r  ggplot-example-4}
p + geom_point(aes(population/10^6, total)) +
  geom_text(aes(population/10^6, total, label = abb))
```

We have successfully added a second layer to the plot. 


### Tinkering with arguments
 
Each geometry function has many arguments other than `aes` and `data`. For example, in the plot we wish to make, the points are larger than the default size. 

```{r ggplot-example-5}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb))
```

`size` is __not__ a mapping: whereas mappings use data from specific observations and need to be inside `aes()`, operations we want to affect all the points the same way do not need to be included inside `aes`.

Now because the points are larger it is hard to see the labels. If we read the help file for `geom_text`, we see the `nudge_x` argument, which moves the text slightly to the right or to the left:

```{r ggplot-example-6}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb), nudge_x = 1.5)
```

This is preferred as it makes it easier to read the text.

## Global versus local aesthetic mappings

In the previous line of code, we define the mapping `aes(population/10^6, total)` twice, once in each geometry. We can avoid this by using a _global_ aesthetic mapping.

If we define a mapping in `ggplot`, all the geometries that are added as layers will default to this mapping. We redefine `p`:

```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
```

and then we can simply write the following code to produce the previous plot:

```{r ggplot-example-7, eval=FALSE}
p + geom_point(size = 3) + 
  geom_text(nudge_x = 1.5)
```

We keep the `size` and `nudge_x` arguments in `geom_point` and `geom_text`, respectively, because we want to only increase the size of points and only nudge the labels. If we put those arguments in `aes` then they would apply to both plots. Also note that the `geom_point` function does not need a `label` argument and therefore ignores that aesthetic.

If necessary, we can override the global mapping by defining a new mapping within each layer. These _local_ definitions override the _global_. Here is an example:

```{r ggplot-example-8}
p + geom_point(size = 3) +  
  geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```

Clearly, the second call to `geom_text` does not use `population` and `total`.


## Scales

First, our desired scales are in log-scale. This is not the default, so this change needs to be added through a _scales_ layer. This particular transformation is so common that __ggplot2__ provides the specialized functions `scale_x_log10` and `scale_y_log10`, which we can use to rewrite the code like this. Because we are in the log-scale now, the _nudge_ must be made smaller.


```{r, eval=FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() 
```


## Labels and titles


To change the labels and add a title, we use the following functions:

```{r ggplot-example-10}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

We are almost there! All we have left to do is add color, a legend, and optional changes to the style.

## Categories as colors

We can change the color of the points using the `col` argument in the `geom_point` function. To facilitate demonstration of new features, we will redefine `p` to be everything except the points layer:

```{r}
p <-  murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

and then test out what happens by adding different calls to `geom_point`. We can make all the points blue by adding the `color` argument:

```{r ggplot-example-11}
p + geom_point(size = 3, color ="blue")
```

This, of course, is not what we want. We want to assign color depending on the geographical region. A nice default behavior of __ggplot2__ is that if we assign a categorical variable to color, it automatically assigns a different color to each category and also adds a legend.

Since the choice of color is determined by a feature of each observation, this is an aesthetic mapping. 

```{r ggplot-example-12}
p + geom_point(aes(col=region), size = 3)
```

The `x` and `y` mappings are inherited from those already defined in `p`, so we do not redefine them. 


## Annotation, shapes, and adjustments

We often want to add shapes or annotation to figures that are not derived directly from the aesthetic mapping; examples include labels, boxes, shaded areas, and lines.

Here we want to add a line that represents the average murder rate for the entire country. Once we determine the per million rate to be $r$, this line is defined by the formula: $y = r x$, with $y$ and $x$ our axes: total murders and population in millions, respectively. In the log-scale this line turns into: $\log(y) = \log(r) + \log(x)$. So in our plot it's a line with slope 1 and intercept $\log(r)$. To compute this value, we use our __dplyr__ skills:

```{r}
r <- murders %>% 
  summarize(rate = sum(total) /  sum(population) * 10^6) %>% 
  pull(rate)
```

The default plots created by __ggplot2__ are already very useful. However, we frequently need to make minor tweaks to the default behavior. Although it is not always obvious how to make these even with the cheat sheet, __ggplot2__ is very flexible.

For example, we can make changes to the legend via the `scale_color_discrete` function. In our plot the word _region_ is capitalized and we can change it like this:

```{r}
p <- p + scale_color_discrete(name = "Region") 
```

## Add-on packages {#add-on-packages}

The power of __ggplot2__ is augmented further due to the availability of add-on packages.
The remaining changes needed to put the finishing touches on our plot require the __ggthemes__ and __ggrepel__ packages.

Many  themes are added by the package __ggthemes__. Among those are the `theme_economist` theme that we used. After installing the package, you can change the style by adding a layer like this:

```{r, eval = FALSE}
library(ggthemes)
p + theme_economist()
```

In our plot, some of the labels fall on top of each other. The add-on package __ggrepel__ includes a geometry that adds labels while ensuring that they don't fall on top of each other. We simply change `geom_text` with `geom_text_repel`.

## Putting it all together

Now that we are done testing, we can write one piece of code that produces our desired plot from scratch. 


```{r final-ggplot-example}
library(ggthemes)
library(ggrepel)



murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
  # geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") + 
  scale_color_discrete(name = "Region") +
  theme_economist()
```


