---
title: "Control structures, exercises"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Control structures, exercises}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## If-else statement

1. load in the Orange dataset
2. Create a new, empty variable
3. Write a for loop to iterate over the rows of the dataset
3. Write an if else statement that classifies trees based on their circumference: 
    <= 65.5 -- small
    >65.5 & <= 161.5 -- medium
    >161.5 large
4. Do the same with ifelse() in one step.

```{r}
data("Orange")

Hmisc::describe(Orange)

```


## For, while loop

1. for each tree (1 to 5), print the ages. 
2. for each tree, print the age, in which they were small. 
3. load esoph dataset. 
4. Using for loop(s), summarize ncontrols by agegp and alcgp, skipping the category alcgp=="120+" & agegp=="25-34".
5. Using a for loop simulate the flip a coin twenty times, keeping track of the individual outcomes (1 = heads, 0 = tails) in a vector that you preallocte.
6. Use a while loop to investigate the number of terms required before the product 1*2*3*4*... reaches above 10 million
7. Use a nested for loop (a for loop inside a for loop) that produces the following matrix, preallocate the matrix with NA values.

```{r}


```


## apply, lapply, etc...

1. define the following dataset:

```{r}
dataset1 <- cbind(observationA = 16:8, observationB = c(20:19, 6:12))
```

2. Calculate row and column means with apply 
3. Use apply to multiply the whole table by 3. 

```{r}



```


```{r}

```

4. create a list with 3 list elements. Print the length of each element. 

```{r}
lll <- list(list(1,3,5,6,1), list(5,2,1), list(6,2,5,7,3,1,9))
```


```{r}


```

