---
title: "Functions, exercises"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions, exercises}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

1. Write a function that divides two numbers. If the divider is zero, it shouldn't try the division, but give a message about it. 

```{r}
my_first_function <- function(num1, num2){
  if (num2==0){
    stop("Not working")
}
  res <- num1/num2
  return(res)
}

```


2. Write a function that changes all vowels of a string to e-s.

```{r}
esperente <- function(c){
  gsub("[aoiu]", "e", c)
}

```


3. Add a checking part, that checks if the input is indeed one character and it has vowels in it. If not, it should give a message. 

```{r}
esperente <- function(c){
  if (!is.character(c)){
    message("Not a character")
  }
  if (length(c)!=1){
    message("Too many elements!")
  }
  if (!grepl("[aoiu]", c)){
    message("The result will be the same!")
  }
  
  gsub("[aoiu]", "e", c)
}

```

4.  Write a function that calculates the variance of a vector.

$$Var(x)=\frac 1{n-1}\sum_{i=1}^{n} (x_{i}-\overline{x})^2$$
where 
$$\overline{x}=\frac{\sum_{i=1}^{n}x_{i}}n$$
```{r}
variance <- function(num){
  
  sum((num-mean(num))^2)/(length(num)-1)
  
}

```


5. Add the respective checking codes. 

```{r}

```

6. Add a warning, if the class of the input vector is not numeric, but can be coarced as one. 

```{r}

```

7. Add errors is the checking codes fail. 

```{r}

```


