---
title: "R Data types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R Data types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Primitive types

## Numeric and integer

```{r, results='asis'}

x <- 1
y <- 2L

cat("R is non-declarative. It means we don't have to declare the type of variables before creating them. \n It also means that the class (type) of the variables will be assigned automatically!")


class(x)
class(y)

cat("Change the class of a variable - coercing \n")
class(as.integer(x))

z <- 1.314
v <- 1.678
v2 <- 2.1
class(z)


print(as.integer(z))
print(as.integer(v))
print(as.integer(v2))

```

## Character

```{r}

a <- "a"
a1 <- 'b'

b <- "The red fox"
class(a)
class(a1)
class(b)


#"Careful!"
c <- "1"
class(c)

```

## Logical 

```{r}

a <- TRUE
b <- FALSE

print(a)
class(a)
#BUT!

c <- "TRUE"

class(c)

print(as.integer(a))
print(as.integer(b))

a <- as.logical(0L)
b <- as.logical(1)
print(a)
print(b)

```


## Operators

### Arithmetic operators

| Operator | Description |
|:---------:|:------------:|
|     +	   |   Addition  |
|     -	   |   Substraction |
|     *	   |   Multiplication |
|     /	   |  Division  |
|     ^	   |   Exponent |
|%%	| Modulus (Remainder from division) |
|%/%	| Integer Division |

### Relational operators

| Operator | Description |
|:---------:|:------------:|
|<	| Less than |
|>	| Greater than |
|<=	| Less than or equal to |
|>=	| Greater than or equal to |
|==	| Equal to |
|!=	| Not equal to |

### Logical operators

| Operator | Description |
|:---------:|:------------:|
|!	| Logical NOT | 
|&	 | Element-wise logical AND |
|&&	| Logical AND |
||	| Element-wise logical OR |
|||	| Logical OR |

### Assignment operators 

| Operator | Description |
|:---------:|:------------:|
|<-, <<-, =	 | Leftwards assignment |
|->, ->>	| Rightwards assignment |

## Operations with primitive types

```{r}

num1 <- 15.12
num2 <- 6.54
num3 <- 1

char1 <- "a"
char2 <- "b"
char3 <- "1"

logic1 <- TRUE
logic2 <- FALSE

num1+num2

num1-num2

num1*num2

num1/num2

num2^2

num1%%num2

num1%/%num2

try(char1+num1)
try(char3+num1)

try(as.numeric(char1)+num1)
try(as.numeric(char3)+num1)

try(logic1+logic2)
try(num1+logic1)

num1<num2
num1!=num2
num1==num2


try(char1<char2) 
# alphabetical order 
try(char1==char2)
try(char1!=char2)
#But!

try(char1>num2)
try(char1<=num3)
try(char3<=num3)



logic1<logic2
logic1!=logic2
logic1==logic2

#Logical operators

TRUE & FALSE
TRUE | FALSE
!TRUE

#Depends on your computer/locale

x1 <- 0.5 - 0.3
x2 <- 0.3 - 0.1
x1 == x2

# should use all.equal instead
all.equal(x1, x2)

#concatenate characters

paste(char1, char2)
paste(char1, char2, sep="_")
paste0(char1, char2)



```

# Simple types (S3 classes) 

## Vectors

Vectors are 1 dimensional collection of data of the <b>same</b> type. 
Indexing in R is always 1 based.

```{r}
#empty vector

a <- vector()
a <- c()

# integer vector

a <- c(1L,2L,3L)
a <- 1:3

typeof(a)
class(a)

#It can handle only one type! It will coerce the data into the most permissive type:
b <- c(1,2,3)

a <- c(1L,"two",3L)
typeof(a)

a <- c(1L,2,3L)
typeof(a)

#vector functions

length(a)
a2 <- c(a,a) #combine vectors
a2

#compare vectors

a==b
a==a
#but!
a==a2

a <- c(TRUE, TRUE, FALSE)
b <- c(TRUE, TRUE, TRUE)

a & b
a && b

a | b
a || b

# subsetting vectors 
# Indexing is 1-based

b <- c(1,2,4, NA)
a <- c(1,2,3)

a[1]
b[c(1,2)]
b[1:2]

a[a>2]
b[b>2]
b[!is.na(b)]
which(!is.na(b))

#named vectors

names(a) <- c("one", "two", "three")

a
a["one"]


paste(a, collapse = ",")

#other useful functions

sum(a)
mean(a)
median(a)
sd(a)

mean(b)
mean(b, na.rm = T)



```

## Factors 

Factors in R are quite unique type. It is very useful in some cases, but in my experience, it can be source of errors if not handled properly. 
It is used for categorical variables. It is internally represented as integers, but has labels. The set of labels are fixed. 

```{r}

a <- factor(c("a", "b", "c", "a"), levels = c("a", "b", "c", "d"))
a

class(a)
typeof(a)

a[2] <- "d"
#but!
a[2] <- "e"

a <- factor(c("1", "2", "3"), levels = c("1", "2", "3", "4"))
b <- factor(c("1", "4", "3"), levels = c("1", "4", "3", "2"))

as.numeric(b)==as.numeric(a)

a==b

sort(a)
sort(b)

as.numeric(as.character(b))==as.numeric(as.character(a))


```
## Date and time

```{r}
x <- as.Date("1971-01-01")
unclass(x)

typeof(x)

attributes(x)

current_date <- Sys.time()
current_date
format(current_date, "%d %b %Y")
current_date <- Sys.Date()

format(current_date-1, "%d %b %Y")

```

## Lists

Similar to vectors, but can take lists as elements. The elements can have different types. 

```{r}

a <- list(c(1, 2),2,3)
a

a <- list(c(1:4), "two", "three")
a

a <- list(list(c(1, 2),2,3,4), "two", "three")
a

#can have names

b <- list(numbers=c(1:4),letters=c("two", "three"))


#referring to an element/subsetting

a[[1]]
a[[1]][[1]]
a[[1]][[1]][1]

b$letters[1]

#functions

length(a)
length(a$letters)


```

## Matrix 

Data format to store values in a matrix (rows and columns). The elements can be primitive types. All element has to be the same type. 

```{r}

letter_mat <- matrix(data=c(rep("a", 3), rep("b", 3)), nrow = 3)

num_mat <- matrix(1:9, nrow=3, ncol=3)

# matrix can have row names but not column names

num_mat*num_mat #element-wise multiplication  
num_mat %*% num_mat #matrix multiplication 
t(num_mat)	#Transpose
# many other matrix specific funtions and operators

#subsetting

letter_mat[1,2]
letter_mat[1,]
letter_mat[1:2]
letter_mat[3]
letter_mat[,1]

#combining
cbind(letter_mat, num_mat)
rbind(letter_mat, letter_mat)

#row and column-wise operations
#matrixStats package - fast!


num_mat <- matrix(runif(100), nrow=10, ncol=10)

rowMeans(num_mat)	
matrixStats::rowMeans2(num_mat)
rowSums(num_mat)	
colMeans(num_mat)
colSums(num_mat)	

```

## Data frame

2D object type. Rows and columns with names. The columns can have different element types. 


```{r}

df <- data.frame(value=c(1,2,3), "First name"=as.factor(c("John", "John", "Peter")), last_name=as.factor(c("Smith", "Black", "Rabbit")), row.names = c("Patient 1", "Patient 2", "Patient 3"))


df[1:2,]

df[c(TRUE, TRUE, FALSE),]

df[c("Patient 1", "Patient 2"),]

df[-3,]

```

