---
title: "S4 classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S4 classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

S4 classes are the perfect examples of the OO programming:
- stricter structure
- they have representations: a list of slots or arguments by name and class
- the slots have to have the correct type
- methods - they also inherit the methods from the classes of the arguments, but inheritence is between S4 classes as well. 
- has to be created with the new constructor
- they have accessor methods as well.

```{r}
setClass("Person", representation(name = "character", age = "numeric"))
setClass("Employee", representation(boss = "Person"), contains = "Person")

person1 <- new("Person", name="Peter", age=34)
boss1 <- new("Employee", boss=person1, name="Lisa", age=36)
```

```{r}

person1
boss1

```
Example class: SummarizedExperiment

Representation:


`RangedSummarizedExperiment` is an extension of `SummarizedExperiment`. It uses `GRanges` to define the genomic regions. 


```{r}
library(SummarizedExperiment)
data("airway_small",  package="BasicR")

se <- airway_small

```
## Accessors

To know all this, look for the help and the vignette. 

```{r}
colData(se)
rowData(se)
#assay(se)
rowRanges(se)
#assays(se)[[1]]
head(assays(se)$counts)

head(metadata(airway_small))

```
You can access each data "natively", and even change it, but you shouldn't do that. 
Why? It is easy to screw it up...


```{r}
se@rowRanges
se@metadata

###don't do this:
se@rowRanges <- se@rowRanges[1:3000]
se
se@rowRanges


```
## Methods 

### subsetting


```{r}
# subsetting is very easy, just like with a data.frame

se <- se[1:2000,]

se <- se[,1:3]

se

```

### Combining

```{r}

rowbinded <- rbind(se, se)
rowbinded
colbinded <- cbind(se, se)
colbinded

```
```{r}
roi <- GRanges(seqnames="X", ranges=99800000:99850000)
se_1 <- subsetByOverlaps(se, roi)



```

## Granges and GRangesList

`GRanges` objects store information about Genomic ranges with chromosome, position and additional information. Offers a wide range of functionality.
`GRangesList` is a "grouped" version of `GRnages`, it is similar to a normal list, but with additional functionality. 
Both of them are able to handle strand information as well.

```{r}
library(GenomicRanges)

gr1 <- GRanges(
    seqnames = "chr2",
    ranges = IRanges(103, 106),
    strand = "+",
    score = 5L, GC = 0.45)
gr2 <- GRanges(
    seqnames = c("chr1", "chr3"),
    ranges = IRanges(c(107, 113), width = 3),
    strand = c("+", "-"),
    score = 3:4, GC = c(0.3, 0.5))
gr3 <- GRanges("chr2:102-107")

gr4 <- makeGRangesFromDataFrame(data.frame(chr=c("chr1", "chr2"), 
                                           start=c(104, 104), end=c(108, 105),
                                           name=c("gene1", "gene2")), keep.extra.columns = T, ignore.strand = T)
grl <- GRangesList("txA" = gr1, "txB" = gr2)
grl

#seqinfo(gr1) <- Seqinfo(genome="hg38")
#can have names for each row


```


### Accessors

```{r}
seqnames(gr1)
seqlevels(gr1)
ranges(gr1)
start(gr1)
end(gr1)
strand(gr1)
width(gr1)
length(gr1)
mcols(gr4)
mcols(gr4)$name

```

### Subsetting


```{r}
#data.frame-like

gr1[1,]
gr4[1, "name"]


# subset function
subset(gr1, strand == "+")

```
### Splitting and combining

```{r}

sp <- split(gr2, 1:2)
split(gr4, ~ name)

c(sp[[1]], sp[[2]])

stack(sp, index.var="name")


```
### Aggregating

```{r}
aggregate(gr2, score ~ strand, mean)

```
### Interval operations

```{r}
flank(gr1, 10, both=T)
shift(gr2, 5)
resize(gr1, 20)
c(gr1, gr3)
 reduce(c(gr1, gr3)) # combines the overlapping ranges

 reduce(c(gr1, gr3), ignore.strand=T)
```
### Comapring 2 GRanges

```{r}
union(gr2, gr4)

#intersect(gr2, gr4)
setdiff(gr2, gr4)

hits <- findOverlaps(gr2, gr4, ignore.strand=TRUE)
countOverlaps(gr2, gr4, ignore.strand=TRUE)
subsetByOverlaps(gr2, gr4)


```



## DeSeq2


```{r}
library("pasilla")
pasCts <- system.file("extdata",
                      "pasilla_gene_counts.tsv",
                      package="pasilla", mustWork=TRUE)
pasAnno <- system.file("extdata",
                       "pasilla_sample_annotation.csv",
                       package="pasilla", mustWork=TRUE)
cts <- as.matrix(read.csv(pasCts,sep="\t",row.names="gene_id"))
coldata <- read.csv(pasAnno, row.names=1)
coldata <- coldata[,c("condition","type")]

rownames(coldata) <- sub("fb", "", rownames(coldata))
cts <- cts[, rownames(coldata)]

library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ condition)
dds
dds <- dds[1:200,]
dds <- DESeq(dds)
res <- results(dds)
res


```

```

