---
title: "Data Cleaning"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---

Please go to Session on the RStudio Menu, and click on 'Restart R and Clear Output'.


# CLEANING DATA 

## Characteristics of a clean dataset

- Follows a tidy data structure
- Remove duplicate rows/values
- Error-free (e.g. free of misspellings)
- Variables should have appropriate data type. e.g. numeric, character, factor etc
- Factors (categorical/ ordinal) should have relevant levels
- Remove incorrect/ non-relevant outliers
- Missing data should be set as NA



## CLEANING DATA Exercise

1. Load the __tidyverse__ libraries

```{r}
library(tidyverse)
```

2. Import `messy_data.csv` data from `/home/rstudio/inst/extdata`

```{r}
messy_data <- read_csv("../inst/extdata/messy_data.csv")
```


3. Use `head`, `summary` and `View` to view the data. Identify reasons why data is not in perfect order for analysis.
```{r}
head(messy_data)
summary(messy_data)
```

Also try `View(messy_data)`
```{r,eval=FALSE}
View(messy_data)
```

4. Remove the column _X_. There are only NAs present in the variable.
```{r}
messy_data %>% pull(X)
messy_data <- messy_data %>% select(-X)

```


5. Remove observations with _Patient.id_ marked as NA. Multiple missing rows need to be removed.
```{r}
tail(messy_data)
messy_data <- messy_data %>% filter(!is.na(Patient.id))
```


6. _Patient.id_ values are not consistent. _TCGA _ as prefix is used in different forms. Use `str_replace_all` and `to_upper` to fix this.
```{r}
messy_data<- messy_data %>% 
  mutate(Patient.id = toupper(messy_data$Patient.id)) %>% 
  mutate(Patient.id =  str_replace_all(messy_data$Patient.id,"tcga", "TCGA")) %>% 
  mutate(Patient.id =  str_replace_all(messy_data$Patient.id,"TCGA ", "TCGA_")) 
```



7. Check categorical variables. 
```{r}
messy_data %>% pull(Gender) %>% table()
messy_data %>% pull(Status) %>% table()
messy_data %>% pull(Censor) %>% table()
messy_data %>% pull(Identifier) %>% table()
messy_data %>% pull(Classification) %>% table()
messy_data %>% pull(Histological.subtype) %>% table()
messy_data %>% pull(MUC1) %>% table()
messy_data %>% pull(TNM) %>% table()

```

8.  _Classification_ is not an informative variables, so remove it.
```{r}
messy_data<-messy_data %>% select(-Classification)
```



9. _MUC1_ has an entry = ???. Mark this as missing. Create a factor variable to ensure levels of factor _MUC1_ are relevant (in ascending order).
```{r}
messy_data<-messy_data %>% mutate(MUC1=ifelse(MUC1=="???",NA,MUC1)) %>%
    mutate(MUC1=factor(MUC1, levels = c("0", "0bis1", "1", "2", "3")))
```



10. Checking numerical variables. 
```{r}

messy_data %>% pull(Survival) %>% summary()
messy_data %>% pull(Survival) %>% hist()

messy_data %>% pull(enzyme.level) %>% summary()
messy_data %>% pull(enzyme.level) %>% hist()

messy_data %>% pull(Age) %>% summary()
messy_data %>% pull(Age) %>% hist()
```


11. _enzyme.level_ has an incorrect outlier. Identify and mark it as missing.
```{r}
messy_data<-messy_data %>% mutate(enzyme.level=ifelse(enzyme.level>600,NA,enzyme.level))
messy_data %>% pull(enzyme.level) %>% hist()
```


### Move PAST OPTIONAL CLEANING STEPS directly to MANIPULATING DATA Exercise below. ###


### <<--- OPTIONAL CLEANING STEPS --->> 

12. _TNM_ contains 3 variables (columns) in 1. Use `substr` to extract the relevant position of string to get T,N and M. Convert them into __integer__ type using `as.integer`.
```{r}
messy_data<-messy_data %>% mutate(T=as.integer(substr(TNM, 2, 2)),
         N=as.integer(substr(TNM, 4, 4)), 
         M=as.integer(substr(TNM, 6, 6)))

messy_data %>% pull(TNM) %>% table()
messy_data %>% pull(T) %>% table()
messy_data %>% pull(N) %>% table()
messy_data %>% pull(M) %>% table()
```

13. Dates are not in __Date__ format.
```{r}

summary(messy_data)
messy_data<-messy_data %>%  
mutate(Date.of.surgery= as.Date(Date.of.surgery, format="%d-%b-%y")) %>%
mutate(Date.of.birth= as.Date(Date.of.birth, format="%d/%m/%Y")) %>%
mutate(Date.of.diagnosis= as.Date(Date.of.diagnosis, format="%d/%m/%Y")) %>%
mutate(Time.of.death= as.Date(Time.of.death, format="%d-%b-%y"))
summary(messy_data)
```

14. _Age_ has an incorrect outlier. Age of 6 is too low. Correct it using the Dates.
```{r}
messy_data %>% pull(Age) %>% summary()
messy_data %>% pull(Age) %>% table()

messy_data<-messy_data %>% mutate(Age=difftime(Date.of.diagnosis,Date.of.birth, units = "days")) %>% 
  mutate(Age=round(Age/365, digits = 0))

messy_data %>% pull(Age) %>% summary()
messy_data %>% pull(Age) %>% table()

```

### <<--- OPTIONAL CLEANING STEPS COMPLETE --->> 


## MANIPULATING DATA Exercise

1\. Focus on _alive_ _females_. `Status` and `Gender` are categorical variables with the respective tags. What is the average and standard deviation of `enzyme.level` for this subset of patients? 

Hint: Use `filter` and `summarize` and use the `na.rm = TRUE` argument when computing the average and standard deviation. You can also filter the NA values using `filter`.
```{r}
```

2\. Using a pipe, assign the average to a numeric variable `ref_avg`. Hint: Use the code similar to above and then `pull`.
```{r}
```

3\. Now report the min and max values for the same group.
```{r}
```

4\. Compute the average and standard deviation for females, but for each Status group separately rather than only _alive_ as in question 1.  Hint: rather than filtering by Status and gender, filter by `Gender` and then use `group_by`.
```{r}
```

5\. We can actually combine both summaries for males and females into one line of code. This is because `group_by` permits us to group by more than one variable. Obtain one big summary table using `group_by(Status, Gender)`.
```{r}
```

6\. For males, compare enzyme.level across Histological subtypes as reported in the `Histological.subtype` variable. Order the resulting table from lowest to highest average `enzyme.level`.
```{r}
```
