---
title: "Multivariable regression"
output:
  html_document:
  highlight: tango
toc: true
toc_float:
  collapsed: true
---


Please go to Session on the RStudio Menu, and click on 'Restart R and Clear Output'.




# MULTIVARIABLE REGRESSION


We will use the same dataset as the previous exercise - The Human Freedom Index from the **openintro** package.

In this lab, you'll be analysing data from the Human Freedom Index reports.
Your aim will be to summarize a few of the relationships within the data both graphically and numerically in order to find which variables can help tell a story about freedom.


## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the **tidyverse** suite of packages.

The data can be found in the **openintro** package. The **broom** package helps us summarize regression results.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(broom)
```


### The data

The data we're working with is in the openintro package and it's called `hfi`, short for Human Freedom Index.

The dataset spans a lot of years, but we are only interested in data from year 2016.
    Filter the data `hfi` data frame for year 2016, select the six variables, and assign the result to a data frame named `hfi_2016`.

```{r}
hfi_2016 <- hfi %>% 
  filter(year == 2016)
```



------------------------------------------------------------------------

## EXERCISES

1\. Copy your model from the previous exercise that uses  `pf_expression_control` to predict Human Freedom or `hf_score`. Using the `tidy` command, what does the slope tell us in the context of the relationship between human freedom and expression control in the country?

2\. Add region to the model from Q1 using `lm(hf_score ~ pf_expression_control + region, data = hfi_2016)`. What do you notice about the slope between human freedom and expression control? How has it changed from the previous model. Do you think region is a confounder, and think about reasons why this might be so? 

3\. Compare the $R^2$ for the 2 models from Q1 and Q2. Is there an increase by adding region? Think about the definition of $R^2$. What does this mean in this context?

4.\ Fit a new model that uses  `ef_money` or monetary measure to predict `hf_score`.  What does the slope tell us in the context of the relationship between human freedom and the economy in the country? 

5\. Again add region to the model from Q4. Compare the slope and $R^2$ with the model from Q4.

6\. Finally fit a model with `ef_money` and `pf_expression_control` as exposures and `hf_score` as outcome.  Compare the slope and $R^2$ from the models from Q1. Could `ef_money` be a confounder? 

7\. Use a linear regression model (and scatter plot) with  `ef_money` as exposure and `pf_expression_control` as outcome, to study whether `ef_money` has an association with `pf_expression_control` as well. This might validate our finding that `ef_money` is a confounder between  `pf_expression_control` as exposure and 
`hf_score` as outcome from Q6.


